\documentclass[10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{microtype}

\title{BELA: A Fully Automated Blastocyst Evaluation Learning Algorithm for
Ploidy Prediction from Time-Lapse Imaging}
\author{[Authors removed for anonymous review]}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Assessing fertilized human embryos is central to the success of in vitro
fertilization (IVF). Existing deep-learning models for ploidy prediction either
rely on subjective embryologist-derived scores or consume entire time-lapse
videos, which limits generalization and clinical deployment. We present BELA
(Blastocyst Evaluation Learning Algorithm), a fully automated two-step model
that predicts embryo ploidy from day-5 time-lapse frames (96--112 hours post
insemination) and maternal age. BELA first extracts 512-dimensional features
from each frame with a pre-trained VGG16 network and feeds these to a multitask
bidirectional LSTM that jointly predicts inner cell mass, trophectoderm,
expansion, and blastocyst score. The resulting model-derived blastocyst score
(MDBS) is then combined with maternal age in a logistic regression to classify
embryos as euploid (EUP) versus aneuploid (ANU) or euploid versus complex
aneuploid (CxA). On the Weill Cornell WCM-Embryoscope test set, BELA achieves
AUCs of $0.76 \pm 0.002$ for EUP vs ANU and $0.826 \pm 0.004$ for EUP vs CxA
when maternal age is included, and an MAE of $1.855 \pm 0.03$ between the
predicted and embryologist-derived blastocyst score. Across external datasets
from IVI Valencia (543 embryos) and IVF Florida (869 embryos), BELA
consistently outperforms a day-5 video baseline ($p < 0.05$). BELA removes the
need for subjective grading and is deployed through a web-based clinical tool
(STORK-V) with a per-embryo inference time of $30 \pm 5$ seconds.
\end{abstract}

\input{intro_relwork}

\input{figures/figures}

\section{Methods}

\subsection{Datasets}
We used four datasets spanning two imaging instruments and three clinics, as
summarized in Table~\ref{tab:datasets} and Figure~\ref{fig:datasets}. The
primary training dataset, WCM-Embryoscope, was collected from the Center for
Reproductive Medicine at Weill Cornell Medicine and comprises 1{,}998
Embryoscope\textregistered{} time-lapse sequences (916 EUP, 494 single
aneuploid [SA], 588 complex aneuploid [CxA]) from 498 patients with an average
of four biopsied embryos per patient. The second Weill Cornell dataset,
WCM-Embryoscope+, contains 841 sequences (410 EUP, 170 SA, 261 CxA) captured
on a newer Embryoscope+\textregistered{} instrument. Two external validation
datasets were obtained from IVI Valencia, Spain (543 embryos: 234 EUP, 309 ANU)
and from IVF Florida (869 embryos: 445 EUP, 202 SA, 222 CxA). The WCM and
Florida datasets include embryologist-derived blastocyst scores (BS); the
Spain dataset does not. All four datasets include maternal age at the time of
oocyte retrieval. Ploidy labels were assigned by preimplantation genetic
testing for aneuploidy (PGT-A) performed with next-generation sequencing at
the respective clinical sites \citep{garciapascual2020ngs}.

\begin{table}[htbp]
\centering
\caption{Characteristics of the four datasets used in this study. EUP:
euploid; SA: single aneuploid; CxA: complex aneuploid; ANU: aggregate
aneuploid (Spain only).}
\label{tab:datasets}
\begin{tabular}{lllrrrrr}
\toprule
Dataset & Instrument & Site & Total & EUP & SA & CxA & ANU \\
\midrule
WCM-Embryoscope  & Embryoscope  & Weill Cornell & 1998 & 916 & 494 & 588 & -- \\
WCM-Embryoscope+ & Embryoscope+ & Weill Cornell &  841 & 410 & 170 & 261 & -- \\
Spain            & Embryoscope  & IVI Valencia  &  543 & 234 & --  & --  & 309 \\
Florida          & Embryoscope+ & IVF Florida   &  869 & 445 & 202 & 222 & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Temporal and spatial processing}
Time-lapse sequences varied substantially in length, start and end times, and
frame rate. To mitigate these biases, we constructed a standardized timeline
at 30-minute intervals from 0 to 150~hpi, yielding 301 standardized frames per
embryo. For each embryo, the image taken closest in time to each standardized
point was assigned to that slot; if no image was available within a 2-hour
window, the slot was filled with a blank frame. We selected a 2-hour boundary
because significant embryonic changes were observed to occur on intervals
longer than 2 hours and because data acquisition rates were typically dense
enough to capture changes within this window. For video classification, we
extracted frames using three parameters (start~hr, end~hr, interval); for
example, a day-2 model uses start~hr~$=24.0$, end~hr~$=48.0$, and
interval~$=2$~hours, yielding 13 frames. Each frame was resized from
$800 \times 800$ to $224 \times 224$. A circle Hough Transform was used for
embryo segmentation to suppress background bias, applied uniformly across all
four datasets. Data augmentation consisted of random horizontal flipping and
random rotations to simulate real-world orientation variability.

\subsection{Feature extraction}
We extracted spatial features from each standardized frame using an ImageNet
pre-trained VGG16 convolutional network from TensorFlow 2.7. The final layer
of the pre-trained network performed average pooling, producing a
512-dimensional feature vector for each frame. We experimented with several
alternative feature extractors but found VGG16 to perform best in our
experiments.

\subsection{Multitask BiLSTM for blastocyst score prediction}
BELA uses a multitask bidirectional LSTM to predict blastocyst score and its
sub-components (ICM, TE, expansion) jointly from the sequence of 512-dimensional
VGG16 features. The architecture consists of one bidirectional LSTM layer
followed by two multi-unit dense layers; each regression target is produced by
a final 1-unit dense head. The loss function is log-cosh and the optimizer is
Adam, with equal loss weights for all prediction tasks. Maternal age is
included as an additional feature in the BiLSTM regression model. Early
stopping with patience $=5$ was used to monitor validation loss on
cross-validation folds. We found that attention mechanisms and stacks of
multiple bidirectional LSTM layers did not significantly improve performance
across tasks ($p > 0.05$). The rationale for multitasking is that ICM, TE, and
expansion scores compose the overall blastocyst score, so a joint model can
share representations across related regression tasks.

\subsection{Ploidy classification}
The second stage of BELA is a logistic regression that takes the model-derived
blastocyst score (MDBS), optionally together with maternal age, as input and
outputs a binary ploidy label. We train logistic regression models for two
tasks: EUP vs ANU and EUP vs CxA, using cross-entropy loss.

\subsection{Training protocol and evaluation}
The WCM-Embryoscope dataset was split 70/30 into training and test partitions.
After applying exclusion criteria (embryo absent from the petri dish,
less than half-visible, or too dim to discern), the usable set contained 1{,}684
of the original 1{,}998 embryos. We used four-fold cross-validation on the
training split to train the BiLSTM regression model and to select
hyperparameters via validation loss. A logistic regression ploidy classifier
was trained on top of each of the four cross-validated BiLSTM models, and
performance metrics were averaged. Evaluation metrics were accuracy,
area under the receiver operating characteristic curve (AUC), precision, and
recall. Between-group comparisons used Student's $t$-test, with Bonferroni
correction applied across tests.

\subsection{Computational resources}
Model training and inference were conducted on an Apple M1 Mac with
TensorFlow Metal, with the deployment model retrained on a BioHPC cluster
using an NVIDIA A40 GPU. Logistic regression models had an average training
time of $2.5 \pm 1.2$ seconds, while BiLSTM models required $30.3 \pm 11$
minutes. The BELA model on the STORK-V platform was trained in 5.23 minutes
on the A40 GPU. Inference for a single embryo on the STORK-V platform took
$30 \pm 5$ seconds (Table~\ref{tab:timing}).

\begin{table}[htbp]
\centering
\caption{Training and inference times for the BELA components.}
\label{tab:timing}
\begin{tabular}{ll}
\toprule
Component & Time \\
\midrule
Logistic regression training        & $2.5 \pm 1.2$ s \\
BiLSTM training                     & $30.3 \pm 11$ min \\
STORK-V deployment training (A40)   & 5.23 min \\
STORK-V single-embryo inference     & $30 \pm 5$ s \\
\bottomrule
\end{tabular}
\end{table}

\section{Results}

\subsection{Training and validation datasets}
We assembled four datasets totalling 4{,}251 embryos: two internal datasets
from Weill Cornell Medicine (WCM-Embryoscope with 1{,}998 sequences and
WCM-Embryoscope+ with 841 sequences) and two external datasets from IVI
Valencia, Spain (543 embryos) and IVF Florida (869 embryos). Time-lapse
sequences typically contained 360--420 frames captured at 0.3-hour intervals
over five days of development. Ground truth ploidy labels were assigned via
PGT-A. Dataset compositions are shown in Figure~\ref{fig:datasets} and
Table~\ref{tab:datasets}.

\subsection{Blastocyst score prediction}
The first stage of BELA predicts the blastocyst score from day-5 time-lapse
frames. On the WCM-Embryoscope test set, the mean absolute error (MAE)
between the model-derived blastocyst score (MDBS) and the ground truth
Embryoscope BS was $1.855 \pm 0.03$. Multitasking improved BS prediction:
multitask BELA achieved an MAE of $1.855 \pm 0.03$ compared with
$1.877 \pm 0.027$ for a non-multitask variant trained only on blastocyst
score, supporting the use of multitask learning
(Figure~\ref{fig:mae_comparisons}, left). We also observed a moderate
correlation (Pearson correlation $>0.7$) between MDBS and embryologist
BS on both training and test partitions of WCM-Embryoscope. Attention
mechanisms and additional bidirectional LSTM layers were explored but did
not significantly improve BS prediction ($p > 0.05$ across all tasks).

\subsection{Ploidy classification on Weill Cornell datasets}
Using the WCM-Embryoscope test set, BELA trained to distinguish EUP from ANU
attained an AUC of $0.66 \pm 0.008$ without maternal age, rising to
$0.76 \pm 0.002$ with maternal age. For the EUP vs CxA task, the AUC was
$0.708 \pm 0.004$ without age and $0.826 \pm 0.004$ with age
(Table~\ref{tab:auc_wcm}, Figure~\ref{fig:auc_wcm}). In all tested scenarios
(with or without age, across test sets, and across prediction tasks), BELA
outperformed a day-5 video baseline trained to predict ploidy directly from
96--112~hpi time-lapse frames ($p < 0.05$). Without maternal age, the
embryologist-annotated BS model surpassed BELA ($p < 0.05$) in all prediction
tasks, except for EUP vs ANU on the WCM-Embryoscope test set. When maternal
age was incorporated, BELA outperformed the embryologist-annotated blastocyst
score model on the WCM-Embryoscope test set ($p < 0.05$), although it still
underperformed the embryologist-annotated model on the WCM-Embryoscope+
dataset.

\begin{table}[htbp]
\centering
\caption{BELA AUC on the WCM-Embryoscope test set, with and without maternal
age, for the two ploidy classification tasks.}
\label{tab:auc_wcm}
\begin{tabular}{lcc}
\toprule
Task & AUC (no age) & AUC (with age) \\
\midrule
EUP vs ANU & $0.66 \pm 0.008$  & $0.76 \pm 0.002$ \\
EUP vs CxA & $0.708 \pm 0.004$ & $0.826 \pm 0.004$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{External validation: Spain}
We compared BELA with the day-5 video baseline on 543 embryos from IVI Valencia,
Spain. Because this dataset provides only EUP/ANU labels, we restricted the
comparison to the EUP vs ANU task. BELA significantly outperformed the day-5
video baseline in both scenarios --- with and without the inclusion of maternal
age ($p < 0.05$). Embryos in the Spain dataset had been artificially hatched on
day 3, which alters later blastocyst morphology and morphokinetics; these
embryos also exhibited bleached zona pellucida and lacked the full expansion
typical of the Weill Cornell training set. To quantify these differences, we
extracted VGG16 feature encodings for each frame between 96 and 112~hpi of each
embryo, averaged the encodings across frames, and reduced them with PCA. The
resulting feature space (Figure~\ref{fig:pca_placeholder}) showed significant
overlap between the two US-based datasets while the Spanish embryos clustered
distinctly. Nevertheless, BELA's performance without maternal age on the Spain
dataset remained comparable to its performance on Weill Cornell, suggesting
general applicability. Models incorporating maternal age showed decreased
performance relative to Weill Cornell, likely reflecting demographic differences
between patient populations.

\subsection{External validation: Florida}
BELA was further assessed on 869 embryos from IVF Florida. BELA's performance
on this dataset declined relative to the other test sets, as did the
performance of logistic regression baselines that use only maternal age and
embryologist-derived blastocyst score. We attribute this decline to weak
correlations between blastocyst score, maternal age, and ploidy within the
Florida dataset. Because maternal age is a strong predictor in all our models,
any decrease in its correlation with ploidy significantly impacts performance.
Moreover, embryologist-derived BS in the Florida dataset was predominantly
centered around a score of 7, reducing the granularity that made BS a strong
predictor in other datasets. This loss of granularity may reflect differing
scoring practices: IVF Florida evaluates blastocysts at 115 and 144~hpi, while
Weill Cornell uses additional earlier time points.

Interestingly, the model-derived blastocyst score from the first BELA module
showed a stronger correlation with ploidy status ($-0.119$) than the
embryologist-derived blastocyst score ($-0.101$) on the Florida dataset,
suggesting that BELA constructs a score mapping that aligns better with
ploidy status than the original Florida scores. To test this hypothesis, an
embryologist at Weill Cornell re-graded the 50 Florida embryos whose MDBS
deviated most from the Florida blastocyst scores. We observed a decrease in
MAE between the MDBS and the re-graded Weill Cornell BS ($4.16$) compared
with the MDBS and the original Florida BS ($5.02$), indicating higher
agreement between the MDBS and the Weill Cornell scoring method
(Figure~\ref{fig:mae_comparisons}, right). This improved mapping helps explain
why BELA with maternal age significantly outperforms the logistic regression
trained on maternal age and embryologist-derived BS on the EUP vs ANU task
for the Florida dataset ($p < 0.05$).

\subsection{STORK-V clinical deployment}
To make BELA usable in clinical practice, we developed a web-based application,
STORK-V, that runs two BELA variants: one trained to distinguish EUP from ANU
and one trained to distinguish EUP from CxA. The required inputs are day-5
time-lapse frames (96--112~hpi) and maternal age. The output includes
probabilities for euploidy, aneuploidy, and complex aneuploidy, along with
intermediate quality scores for further analysis. Single-embryo inference on
STORK-V takes $30 \pm 5$ seconds (Table~\ref{tab:timing}), and the deployment
model was trained in 5.23 minutes on an NVIDIA A40 GPU, demonstrating the
practicality of running BELA on accessible hardware.

\section{Discussion}
BELA is a fully automated ploidy prediction model that surpasses traditional
IVF embryo classification methods relying on late-stage development or
single-modality (image- or video-only) inputs. By predicting a blastocyst
score as an intermediate representation and combining it with maternal age,
BELA achieves performance competitive with logistic regression models trained
on embryologist-annotated BS and significantly exceeds the performance of
day-5 video models trained directly on time-lapse inputs. Remarkably, BELA
requires only time-lapse frames from 96 to 112~hpi and maternal age to
predict ploidy, making it straightforward to integrate into existing clinical
workflows. BELA also offers a degree of explainability: embryologists can
examine the model-derived blastocyst score and the multitask scores for ICM,
TE, and expansion to understand the rationale behind a specific classification.

Several limitations warrant discussion. First, video classification models
demand substantial training data; with approximately 2{,}000 time-lapse
sequences available, we could not fully explore larger video classification
architectures such as 3D ConvNets or two-stream inflated convolutional
networks. Second, despite trying multiple feature extractors, the ImageNet
pre-trained VGG16 architecture performed best in our setting; alternative
backbones may extract more information from earlier developmental stages.
Third, we did not have access to potentially relevant maternal features such
as hormone levels at oogenesis or broader clinical demographics, which could
improve ploidy prediction. Fourth, using blastocyst scores as intermediary
labels introduces dependence on a manually curated signal that is subject to
intra- and inter-observer variability
\citep{paternot2009intra,paternot2011multicentre,hammond2020freeze,richardson2015grading};
nonetheless, we show that BS remains predictive of ploidy and therefore is
useful as an intermediate proxy. Fifth, differing inclusion and exclusion
criteria between datasets may partially explain the performance differences
across test sets. Sixth, we developed BELA to exclude mosaic embryos; mosaic
embryos with high implantation potential could therefore be misclassified.
Finally, PGT-A results themselves are subject to variability across
laboratories and platforms \citep{maxwell2016miscarriage,friedenthal2020ngs},
and BELA is applicable only to clinics equipped with time-lapse imaging.

Despite these limitations, BELA provides a non-invasive, cost-effective, and
standardized supplement to PGT-A. By preselecting embryos with a higher
probability of being euploid, it can reduce the number of unnecessary
biopsies and associated risks. By de-prioritizing embryos likely to be
complex aneuploid, it can provide early insight into embryos likely to fail
implantation --- a particularly important capability for patients with a
small number of embryos. Future work includes integration of additional
maternal clinical data, handling of mosaic embryos, and deployment in
prospective clinical studies.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
