\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{microtype}
\usepackage{hyperref}

\title{A hippocampus-accumbens code guides goal-directed appetitive behavior}
\author{Anonymous authors}
\affil{Target venue: Nature Neuroscience}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Neurons in dorsal hippocampus (dHPC) encode a rich repertoire of task-relevant environmental features, while downstream structures such as the nucleus accumbens (NAc) translate this information into motivationally appropriate action. Yet the content and the causal role of the dHPC-to-NAc information stream during goal-directed behavior have remained poorly defined. Here, we combine dual-color two-photon calcium imaging with projection-targeted optogenetic stimulation to characterize dHPC neurons that project to NAc (dHPC$\rightarrow$NAc) in mice performing a hidden-reward treadmill task. Across 19 imaging sessions in 6 mice, we recorded 5{,}372 neurons, of which 444 were identified as NAc-projecting by retrograde mCherry labeling. NAc-projecting neurons were enriched for place cells (169/444, 38\%) relative to non-labeled neurons (1{,}581/4{,}928, 32\%; $\chi^2$, $p=0.012$), carried more spatial information, and preferentially overrepresented the reward zone during high-success sessions. They were also enriched for speed-inhibited cells (21\% vs 15\%, $p=0.00022$) and for appetitive-licking-excited cells (7.4\% vs 3.8\%, $p<0.001$). A generalized linear model revealed a substantially higher proportion of conjunctive coders of position, velocity and licking in dHPC$\rightarrow$NAc neurons (44\% vs 19\%, $p<0.001$), and conjunctive neurons allowed more accurate reward-zone decoding. Optogenetic activation of dHPC terminals in NAc evoked mouth movement ($t(3)=7.485$, $p=0.00494$) and deceleration ($t(3)=-3.551$, $p=0.0381$) in ChR2 but not EYFP mice, linking the projection causally to the approach-consummatory transition. Together, these results describe the dHPC-NAc pathway as a specialized channel for reward-anchored spatial and behavioral-state information that drives goal-directed appetitive action.
\end{abstract}

\input{intro_relwork.tex}

\section{Results}

\subsection{Mice acquire a spatial reward task with anticipatory licking}

To probe projection-specific dHPC coding during goal-directed behavior, we first established a head-fixed spatial reward task in which mice ran on a cue-enriched self-propelled treadmill (belt length 360 cm, textured into six distinguishable zones) and obtained a liquid reward by licking within a hidden 30-cm reward zone (Fig.~\ref{fig:schematic}). We trained a cohort of 18 mice over five consecutive days. Across training, the number of rewarded laps per session (repeated-measures ANOVA, $F(4)=6.344$, GG-correction), anticipatory licking ($F(4)=3.803$) and reward licking ($F(4)=4.276$, GG-correction) all increased significantly, consistent with the acquisition of a spatial anticipatory strategy. Supplementary behavioral metrics---running velocity, lap count, lick count and rewarded-lap ratio---also rose over training (all $p<0.05$). These patterns are consistent with previously reported behavioral profiles on treadmill reward tasks, in which mice progressively concentrate licking near remembered goal locations \citep{Hok2007Goal,Mamad2017PlaceField}, providing a well-defined window for interrogating the hippocampal code that supports this behavior.

\subsection{NAc-projecting dHPC neurons are enriched for place cells with higher spatial information}

\textit{Context.} We next asked whether projection-defined dHPC neurons differ from the surrounding local population in their place-coding properties. To address this, we combined retrograde AAVrg-pgk-Cre injections in NAc \citep{Tervo2016AAVretro} with Cre-dependent AAV5-hSyn-DIO-mCherry in dHPC, in Thy1-GCaMP6s mice \citep{Dana2014Thy1GCaMP6,Chen2013GCaMP6}, yielding simultaneous pan-neuronal green (GCaMP6s) and projection-specific red (mCherry) signals in the same imaging field \citep{Pnevmatikakis2017NoRMCorre,Giovannucci2019CaImAn}.

\textit{Approach.} Across 19 imaging sessions in 6 mice, we recorded a total of 5{,}372 neurons, 444 of which co-expressed mCherry and were classified as dHPC$\rightarrow$NAc. Place cells were defined by comparing each neuron's spatial information content against its own shuffled distribution.

\textit{Finding.} The proportion of place cells was significantly higher in dHPC$\rightarrow$NAc neurons (169/444; 38\%) than in dHPC$-$ neurons (1{,}581/4{,}928; 32\%; $\chi^2$, $p=0.012$; Fig.~\ref{fig:place_conj}). Within place cells, dHPC$\rightarrow$NAc cells had higher spatial information rates (Welch's $t(186.55)=4.770$, $p<0.001$), higher sparsity (Welch's $t(218.79)=3.657$, $p<0.001$), higher within-place-field reliability (Welch's $t(203.46)=2.479$, $p=0.014$), and higher relative in-field activity (Welch's $t(190.32)=2.884$, $p=0.0044$). Place-field widths also differed between populations (Kolmogorov-Smirnov, $D=0.116$, $p=0.0299$), with projection-specific quartiles of 36.5/58.9/69.0 cm compared to 35.6/51.7/69.0 cm for dHPC$-$.

\textit{Interpretation.} The NAc-projecting subpopulation therefore forms a higher-fidelity place-coding compartment within dHPC, consistent with the idea that reward-oriented downstream targets are read out from a subset of CA1 neurons with particularly informative spatial representations \citep{Xiao2020Conjunctive,Ciocchi2015Selective}.

\subsection{dHPC$\rightarrow$NAc place fields overrepresent the reward zone during successful behavior}

\textit{Context.} If the dHPC-NAc channel specifically supports goal-directed behavior, place fields in this population should be biased toward reward-relevant locations, particularly on behaviorally successful sessions \citep{Hok2007Goal,Gauthier2018Reward,Sosa2018Reward,Sato2020Overrepresentation}.

\textit{Approach.} We compared place-field density near the reward zone (reward plus preceding anticipation zone) to elsewhere on the belt, separately for each population and for high- vs. low-success trials, defined by lick precision and reward dispensation.

\textit{Finding.} Place-field starts and ends were both biased toward the reward zone more strongly in dHPC$\rightarrow$NAc neurons (PF start: $p=0.033$; PF end: $p=0.036$, $\chi^2$). Splitting sessions by behavioral performance, we found significantly more dHPC$\rightarrow$NAc place cells near the reward zone in high-success than low-success sessions (Welch's $t$, $p=0.0036$); dHPC$-$ neurons showed only a trend in the same direction ($p=0.051$). A two-way ANOVA confirmed a main effect of success ($F_{\text{success}}(1,1)=54.918$, $p<0.001$) with no main effect of projection ($F_{\text{proj}}(1,1)=0.958$, $p=0.338$) and a weak interaction ($F_{\text{int}}(1,1)=2.969$, $p=0.098$); post-hoc Welch's tests with Bonferroni correction yielded $t_{\text{dHPC}-}(3.561)=3.698$ and $t_{\text{dHPC}\rightarrow\text{NAc}}(3.479)=8.671$. Reward-zone overrepresentation further correlated with behavioral success rate across sessions ($r_{\text{dHPC}-}(15)=0.622$, $p=0.0107$; overall $r=0.62$). A linear decoder trained on calcium activity identified the reward-anticipation zone significantly more accurately from dHPC$\rightarrow$NAc than dHPC$-$ activity within the same sessions (Wilcoxon, $p=0.0195$; $W(9)=5.0$, $n=10$ sessions).

\textit{Interpretation.} The dHPC-NAc channel is thus not merely enriched for place cells, but enriched for \emph{reward-anchored} place cells whose overrepresentation scales with behavioral success, a property not observed to the same extent in the surrounding population.

\subsection{dHPC$\rightarrow$NAc neurons are biased toward speed-inhibited coding}

\textit{Context.} Hippocampal and parahippocampal neurons encode not only position but also running speed, a signal that can be either positively or negatively correlated with movement. Which of these speed codes is preferentially routed to NAc is unknown.

\textit{Approach.} We averaged each neuron's calcium activity across 1 cm/s velocity bins (2--30 cm/s) and performed linear regression; significance was assessed at $p<0.05$ after Benjamini-Hochberg FDR correction.

\textit{Finding.} Approximately 15\% of neurons were positively speed-modulated, with comparable proportions in dHPC$-$ and dHPC$\rightarrow$NAc populations (13\% vs 16\%, $\chi^2(1,5372)=2.565$, $p=0.109$; Fig.~\ref{fig:speed}). In contrast, speed-inhibited neurons were substantially more frequent in the NAc-projecting population (15\% vs 21\%, $\chi^2(1,5372)=13.66$, $p=0.00022$). Representative regressions for speed-excited and speed-inhibited neurons yielded, respectively, slope = $7.43\times10^{-4}$ (intercept $-2.097\times10^{-3}$, $r=0.937$, $p=0.0058$) and slope = $-1.0437\times10^{-4}$ (intercept $2.814\times10^{-3}$, $r=-0.933$, $p=0.0065$).

\textit{Interpretation.} dHPC$\rightarrow$NAc neurons therefore preferentially report periods of \emph{low} locomotion speed---exactly the behavioral state most likely to coincide with the approach and consumption of reward. This asymmetry converts an otherwise symmetric speed code into a state signal aligned with the transition from running to appetitive action.

\subsection{Appetitive-licking-excited neurons are overrepresented in dHPC$\rightarrow$NAc}

\textit{Context.} Decelerations in our task co-occur with increased appetitive licking (Fig.~1B-D); we therefore asked whether licking itself is differentially represented in projection-defined dHPC populations.

\textit{Approach.} For each neuron, we compared pre- and post-lick calcium activity around every appetitive lick event and classified cells as lick-excited or lick-inhibited based on significant pre-post differences.

\textit{Finding.} In total, 1{,}268 neurons (24\% of the imaged population) were significantly modulated by appetitive licking. While the proportion of lick-inhibited neurons was similar across populations (19.7\% in dHPC$-$ vs 17.1\% in dHPC$\rightarrow$NAc; $\chi^2(1,5372)=1.626$, $p=0.202$; Fig.~\ref{fig:lick}), lick-excited neurons were nearly twice as common in dHPC$\rightarrow$NAc as in dHPC$-$ (7.4\% vs 3.8\%; $\chi^2(1,5372)=13.018$, $p=0.00031$). A two-way mixed ANOVA showed a main effect of projection ($F_{\text{proj}}(1,5370)=43.779$, $p<0.001$) and a projection $\times$ lick-timing interaction ($F_{\text{int}}(1,5370)=7.073$, $p=0.0079$), with post-hoc tests indicating that the pre-vs-post calcium difference was significant only in dHPC$\rightarrow$NAc neurons ($t_{\text{dHPC}\rightarrow\text{NAc}}(443)=2.470$, $p=0.0277$) and not in dHPC$-$ ($t(4927)=0.871$, $p=0.768$). Lick modulation (average of each trial's pre-post difference) was higher in dHPC$\rightarrow$NAc neurons (Welch's $t(496.11)=2.379$, $p=0.0177$), as was a lick-index measure (Welch's $t(454.19)=2.729$, $p=0.0066$).

\textit{Interpretation.} The NAc-projecting channel thus preferentially carries signals tied to the consummatory component of reward-seeking behavior, complementing its bias toward speed-inhibited and reward-zone-biased coding.

\subsection{Optogenetic activation of dHPC terminals in NAc elicits mouth movement and deceleration}

\textit{Context.} The correlational enrichment of licking- and deceleration-related coding in dHPC$\rightarrow$NAc neurons raises the possibility that this pathway is not merely observing, but actively driving, the approach-consummatory transition.

\textit{Approach.} We injected CaMKII$\alpha$-driven ChR2 ($n=4$) or EYFP ($n=3$) control virus into dHPC and implanted bilateral optical fibers over NAc. After habituation, head-fixed mice received 5 mW, 473 nm, 20 Hz (5 ms pulse) stimulation of up to 10 s whenever they entered a hidden light-stimulation zone.

\textit{Finding.} Shortly after stimulation onset, ChR2-expressing mice reliably increased mouth movement for up to two seconds after stimulation (paired $t(3)=7.485$, $p_{\text{ChR2}}=0.00494$; Fig.~\ref{fig:opto}), while EYFP controls showed no effect ($t(2)=1.353$, $p_{\text{EYFP}}=0.309$; also reported as $p_{\text{EYFP}}=0.617$ in the paired-test variant). The same stimulation significantly decelerated running on the treadmill in ChR2 animals ($t(3)=-3.551$, $p=0.0381$), with no change in EYFP controls ($t(2)=-1.263$, $p=0.334$).

\textit{Interpretation.} Acute activation of the dHPC-NAc projection is therefore sufficient to produce exactly the behavioral state that the projection most strongly represents during natural task performance: slowed locomotion coupled with initiation of mouth movement. This causal match between correlational coding and optogenetic effect provides direct evidence that dHPC$\rightarrow$NAc projections are a functionally relevant substrate for appetitive action.

\subsection{A generalized linear model reveals enhanced conjunctive coding in dHPC$\rightarrow$NAc}

\textit{Context.} The preceding analyses each isolate a single behavioral variable. We next asked whether the coding differences across projection populations reflect richer, \emph{conjunctive} encoding of multiple variables in dHPC$\rightarrow$NAc neurons.

\textit{Approach.} For each neuron, we fit a generalized linear model (GLM) predicting calcium activity from position, velocity and appetitive licking. Feature importance was quantified via standardized coefficients (H2O \texttt{varimp}), and significance was assessed by comparing each full model against $3\times100$ shuffled surrogates in which one feature was randomized at a time. A feature was deemed significant if its shuffle reduced explained variance in more than 95\% of the surrogate models.

\textit{Finding.} GLMs explained on average close to 40\% of variance in held-out data, with dHPC$\rightarrow$NAc neurons showing slightly lower $R^2$ (Welch's $t(615.15)=2.147$, $p=0.032$) but higher standardized feature importance for position (Welch's $t(2508.09)=2.162$, $p=0.031$) and licking (Welch's $t(1879.75)=9.328$, $p<0.001$), with no difference for velocity (Welch's $t(2609.15)=0.881$, $p=0.378$). The mean reduction in $R^2$ upon shuffling was larger in dHPC$\rightarrow$NAc for position (Welch's $t(523.23)=5.984$, $p<0.001$) and velocity (Welch's $t(497.69)=7.704$, $p<0.001$) but not licking (Welch's $t(552.63)=0.363$, $p=0.717$). Cells encoding space, velocity, or licking individually were each overrepresented in dHPC$\rightarrow$NAc compared to dHPC$-$ (space: $p<0.001$; velocity: $p<0.001$; licking: $p=0.0015$; all $\chi^2$). Crucially, all pairwise dual-coding combinations and the triple-conjunctive class were also overrepresented in dHPC$\rightarrow$NAc (all $p<0.001$), yielding a striking difference at the population level: 44\% of dHPC$\rightarrow$NAc neurons qualified as conjunctive coders, versus 19\% of dHPC$-$ ($p<0.001$; Fig.~\ref{fig:place_conj}). Non-coding neurons were correspondingly overrepresented in dHPC$-$ ($\chi^2(1,5372)=35.382$), single-coding neurons were comparably distributed ($\chi^2(1,5372)=0.0057$), and dual- and triple-coding neurons were overrepresented in dHPC$\rightarrow$NAc ($\chi^2(1,5372)=61.336$ and $\chi^2(1,5372)=30.447$, respectively). Finally, a linear classifier decoding reward-zone presence achieved higher accuracy from conjunctive-coding than non-conjunctive-coding neurons (Wilcoxon, $W(18)=14.0$, $p<0.001$).

\textit{Interpretation.} The dHPC-NAc channel thus encodes the joint structure of position, speed, and consummatory action, rather than a disjoint set of behavioral variables. Conjunctive coding is a candidate mechanism by which downstream NAc readouts could recover reward-relevant context directly from dHPC spike patterns \citep{Xiao2020Conjunctive,Sato2020Overrepresentation}.

\input{figures/figures.tex}

\section{Discussion}

We combined projection-targeted dual-color two-photon imaging with optogenetic manipulation to describe, at population scale, what dHPC neurons projecting to NAc encode during a goal-directed appetitive task. Five inter-related findings define this pathway as a specialized information channel. First, dHPC$\rightarrow$NAc neurons are enriched for place cells and carry more spatial information per spike. Second, their place fields preferentially tile the reward zone, and this tiling tracks behavioral success. Third, they are selectively enriched for speed-inhibited coding, a representation aligned with the approach phase of reward seeking. Fourth, they are selectively enriched for appetitive-licking-excited neurons, encoding the consummatory phase. Fifth, a GLM reveals that all of these biases aggregate into a dramatic overrepresentation of conjunctive coders (44\% vs 19\%), which in turn support more accurate reward-zone decoding. Optogenetic activation of dHPC terminals in NAc reproduces the behavioral states that these neurons most strongly encode, linking representation to action.

Our results extend, and in important ways refine, prior accounts of hippocampal reward representation. Work identifying dedicated reward-responsive populations in hippocampus \citep{Gauthier2018Reward,Sosa2018Reward} and place-field overrepresentation near goals \citep{Hok2007Goal,Mamad2017PlaceField,Sato2020Overrepresentation,Kaufman2020LocusCoeruleus} has, to date, been carried out largely in bulk CA1 populations. We show that several of these features are concentrated in a projection-defined subset that is functionally specialized for NAc communication. This projection specificity may also reconcile apparent inconsistencies across studies: tasks that report no place-cell sensitivity to spatial goal value \citep{Duvelle2019PlaceCells} may be sampling predominantly non-projection populations, in which the bias is attenuated.

The prominence of conjunctive space-velocity-lick coding in dHPC$\rightarrow$NAc neurons is consistent with theoretical proposals that conjunctive codes provide a substrate for downstream, context-sensitive readouts \citep{Xiao2020Conjunctive}. Because NAc itself integrates cortical, thalamic and dopaminergic inputs to adjudicate action selection \citep{Mogenson1980Motivation,Britt2012Synaptic,Jung2024Dopamine}, receiving a conjunctive signal from dHPC---rather than separate position, speed and lick signals---may enable NAc principal neurons to detect reward-relevant states directly from hippocampal spike patterns. This view aligns with recent demonstrations that dHPC-NAc input can drive reinforcement via accumbal dynorphin neurons \citep{Ibrahim2024DorsalHPC} and that NAc participates in a dopamine-dependent memory module for goal-directed navigation \citep{Jung2024Dopamine}, and it dovetails with evidence that ventral hippocampal and prefrontal NAc inputs cooperate to control engagement \citep{Iyer2025Engagement}.

Our optogenetic results add a causal dimension. Acute activation of dHPC terminals in NAc reliably elicited mouth movement and deceleration, two defining features of the approach-consummatory transition and precisely the features that dHPC$\rightarrow$NAc neurons over-encoded in the same animals. This tight match between population coding and behavioral effect is unusual and suggests that dHPC$\rightarrow$NAc activity does not simply reflect reward-related states in parallel to action selection, but is part of the circuitry that produces them. Related ventral hippocampus manipulations have been shown to bias valence, contextual reward memory and engagement \citep{Ciocchi2015Selective,Nguyen2024Cholecystokinin,Iyer2025Engagement,Bakoyiannis2023Pathway}; the dorsal branch, as described here, appears to sit closer to the action-initiation end of this spectrum.

Several limitations merit explicit acknowledgement. Our paradigm is head-fixed, which restricts the range of naturalistic behaviors we can probe and precludes full 3D reward geometries. We imaged only dHPC; a complete understanding of the dHPC$\rightarrow$NAc channel will require simultaneous or closely coordinated recording in NAc to test how conjunctive dHPC patterns are read out by downstream striatal circuits. Our optogenetic sample sizes are small ($n=4$ ChR2, $n=3$ EYFP), and the stimulation is phasic and non-physiological; future experiments using closed-loop, physiologically patterned activation, and targeted inhibition, will be important to establish necessity in addition to sufficiency. Finally, we did not perturb or record dopaminergic inputs to NAc; the interaction between dHPC conjunctive coding and dopaminergic teaching signals \citep{Mamad2017PlaceField,Jung2024Dopamine} remains an open question.

More broadly, our findings suggest a view in which dHPC does not broadcast a uniform cognitive map to all downstream targets, but instead routes a behaviorally-specialized subset of its code to each. The NAc-bound subset appears to be optimized for reporting---and enabling---the moment-to-moment transition from navigation toward a remembered goal to its consummation. Extending this projection-centric view to other dHPC targets, and to pathological states in which reward-guided action is disrupted, is a natural next step.

\section{Methods}

\subsection{Animals and viral strategy}

All experiments were performed according to Directive 2010/63/EU and were approved by the animal care committee of North Rhine-Westphalia, Germany. C57Bl/6 ($n=16$) and Thy1-GCaMP6s (GP4.3; Jackson Laboratory) mice ($n=6$) were bred under specific pathogen-free conditions. Food restriction targeted 80\% of baseline daily pellet intake over 24 h and produced a 10--20\% weight reduction prior to training.

For projection-specific labeling, retrograde AAVrg-pgk-Cre (Addgene \#24593, $1\times10^{13}$ vg/ml) was injected bilaterally into NAc and Cre-dependent AAV5-hSyn-DIO-mCherry ($1.1\times10^{13}$ vg/ml, Addgene \#50459) into ipsilateral dorsal hippocampus (3.38 mm AP, $\pm 2.5$ mm ML, 1.8 mm DV at 10$^\circ$) to drive mCherry expression exclusively in NAc-projecting dHPC neurons \citep{Tervo2016AAVretro}. For optogenetic experiments, AAV2-CaMKII-hChR2(H134R)-EYFP-WPRE ($4\times10^{12}$ TU, UNC \#AV4381E) or control rAAV2-CaMKII-EYFP ($4.3\times10^{12}$ TU, UNC \#AV6650) was injected bilaterally into dHPC; fiber-optic cannulae (TFC\_200/230-0.37, Doric Lenses) were implanted over NAc (+1.3 mm AP, $\pm 1$ mm ML, $-4.9$ mm DV).

\subsection{Behavioral task}

Mice ran on a self-propelled treadmill (Luigs \& Neumann) with a 7 cm $\times$ 360 cm textile belt, textured into six distinguishable zones. A hidden 30-cm reward zone was placed between the horizontal glue stripes and the vertical tape stripes. Rewards were delivered at a spout upon licking within the reward zone. Successful laps were defined as laps in which mice distributed at least 50\% of their licks within reward and anticipation zones and received reward; high-success trials were those with at least 50\% successful laps. Mice were trained for five consecutive days before imaging.

\subsection{Two-photon imaging}

Imaging was performed using an upright LaVision BioTec TrimScope II resonant-scanning microscope with a Chameleon Ultra II Ti:sapphire laser at 920 nm for GCaMP6s and a Spectra Physics HighQ-2-IR laser at 1045 nm for mCherry, through a 16$\times$ water-immersion objective (Nikon N16XLWD-PF). Dual-channel stacks (512$\times$512 or 1024$\times$1024 px, 350--850 $\mu$m field of view) were acquired at 15.2 or 30.5 Hz. Data were motion-corrected with NoRMCorre piecewise-rigid registration \citep{Pnevmatikakis2017NoRMCorre} and spatio-temporal components were extracted with constrained non-negative matrix factorization via CaImAn v1.6.2 \citep{Giovannucci2019CaImAn}. mCherry-positive components were identified by overlaying footprints on the motion-corrected red channel with a dynamic threshold set to visually optimize signal discovery.

\subsection{Place field and spatial coding analysis}

Belt positions were discretized into 45 bins of 8 cm, with only time points at velocity $>2$ cm/s retained. Spatial information (bits/s) was computed as in \citet{Skaggs1993Information} and multiplied by the average firing rate during movement. Place cells were defined by comparing each neuron's spatial information to its shuffled distribution at $p<0.05$. Place fields were obtained by averaging deconvolved activity per bin at velocity $>2$ cm/s, replicating the trace to handle circularity, and applying a Savitzky-Golay filter (window 5 frames, order 2). Place field centers were computed via polar-coordinate centroids. ``Place fields near reward zone'' denotes place cells with centers in either the 30-cm anticipation zone or the 30-cm reward zone.

\subsection{Reward-zone decoding}

Non-zero deconvolved event amplitudes were quantile-normalized and data were restricted to velocity $>2$ cm/s and downsampled to 1 Hz. Binary labels marked whether the current position was within the reward or anticipation zone. Linear decoders were trained per session; only sessions with $>10$ projection neurons were included in across-population comparisons.

\subsection{Speed and lick coding}

Velocity modulation was assessed by linear regression of average deconvolved event rate against 1 cm/s velocity bins from 2--30 cm/s (SciPy \texttt{linregress}) with Benjamini-Hochberg FDR correction at $p<0.05$. Appetitive lick events were defined from down-sampled lick traces (3 Hz), and pre-post-lick calcium activity was compared per neuron to classify lick-excited and lick-inhibited cells.

\subsection{Generalized linear model}

For each neuron, a generalized linear model (H2O implementation) was fit to predict calcium activity from (i) one-hot spatial bin indicators, (ii) normalized velocity, and (iii) appetitive lick bouts, each replicated at time shifts from $-3$ to $+3$ s at 3 Hz to capture anticipatory and delayed activity. Feature importance was assessed by comparing full-model $R^2$ against $3\times100$ shuffled surrogates, with significance declared when a feature's shuffle reduced $R^2$ in more than 95\% of surrogates.

\subsection{Optogenetics}

Following habituation, head-fixed mice received stimulation via a fiber-coupled 473 nm LuxX 473-80 diode laser (Omicron-Laserage) at 5 mW. Trains of 5-ms pulses at 20 Hz were delivered for up to 10 s or until the mouse passed beyond the stimulation zone. Mouth motion was quantified from IR video motion energy, and velocity was extracted from treadmill rotary sensors. Effects were tested by paired $t$-tests comparing pre- and during-stimulation windows.

\subsection{Statistics}

All data are reported as mean $\pm$ SEM unless stated otherwise; significance threshold was $p<0.05$. Proportions were compared with $\chi^2$ tests, continuous distributions with Welch's $t$-tests, and paired optogenetic comparisons with paired $t$-tests; multi-factor designs used two-way ANOVA with Bonferroni-corrected post-hoc tests. Exact test statistics are reported with each result.

\bibliographystyle{unsrtnat}
\bibliography{references}

\end{document}
