\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{microtype}

\title{A SUMO E3 ligase promotes long non-coding RNA transcription to
regulate small RNA-directed DNA elimination}
\author{[Author names withheld]}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Small RNAs silence complementary chromatin regions through interactions with
nascent long non-coding RNAs (lncRNAs), yet in most systems the factors that
enable this ``paradoxical'' transcription are poorly understood. In the
ciliated protozoan \emph{Tetrahymena thermophila}, programmed DNA elimination
is guided by scan RNAs (scnRNAs) that are selected for germline-limited
sequences through target-directed small RNA degradation (TDSD) in the
parental macronucleus (MAC), a process that requires parental-MAC long
non-coding RNAs (pMAC-lncRNAs). Here we identify Ema2 (TTHERM\_00113330), a
conjugation-specific SUMO E3 ligase with an atypical SP-RING domain, as a
key driver of pMAC-lncRNA transcription. Ema2 is expressed exclusively
during conjugation and localizes first to the parental MAC and later to
the new MAC. In its absence, high molecular weight SUMOylated proteins
(mainly $>\!200$~kDa) are reduced to approximately 50\% of wild-type levels,
MDS-complementary scnRNAs fail to undergo TDSD, pMAC-lncRNAs do not
accumulate at three tested IES boundaries, H3K27me3 is lost from the
parental MAC, DNA elimination is partially blocked, and no viable sexual
progeny are produced. In vitro pull-down and mass spectrometry experiments
identify Ubc9 as a direct binding partner and the transcription elongation
factor Spt6 as the major Ema2-dependent SUMOylation target. Cell
fractionation shows that Ema2 promotes the chromatin association of both
Spt6 and RNA polymerase~II at the mid-conjugation stage, in an RNA-independent
manner. A SUMOylation-defective Spt6 mutant (Spt6-C-KR) nevertheless
preserves chromatin association of Spt6 and RNAPII and pMAC-lncRNA
accumulation, while still producing a mild DNA elimination defect,
indicating that Ema2 acts through additional SUMOylated substrates. Our
results place Ema2-directed SUMOylation upstream of lncRNA transcription and
reveal a distinctive route by which a SUMO pathway licenses the
genome-to-small-RNA communication required for programmed DNA elimination.
\end{abstract}

\input{intro_relwork.tex}

\section{Results}

\subsection{Ema2 is exclusively expressed during conjugation and localizes
to the MAC}

As part of a systematic survey of genes upregulated during conjugation
\citep{loidl2021meiosis,miao2009microarray}, we examined the function of
EMA2 (TTHERM\_00113330). Using microarray expression data
\citep{miao2009microarray}, we found that EMA2 mRNA is expressed
exclusively during conjugation (Figure~\ref{fig:timeline}). To localize the
Ema2 protein, we tagged the endogenous \emph{EMA2} locus with a C-terminal
HA epitope (see Materials and Methods). Ema2-HA was undetectable in
vegetatively growing cells and first appeared in the parental MAC at 3 and
6~hours post-mixing (hpm) of two mating types. At the onset of new MAC
development (8~hpm), Ema2-HA relocated from the parental MAC to the new
MAC, and by 12 and 14~hpm the signal had faded
(Figure~\ref{fig:timeline}). This conjugation-specific expression and the
switch from parental to new MAC are reminiscent of other factors involved
in programmed DNA elimination, including the Piwi protein Twi1 that is
loaded by scnRNAs \citep{mochizuki2002twi1,noto2010giw1} and PRC2
components \citep{liu2007rnai}.

\subsection{Ema2 is required for completing programmed DNA elimination}

To test whether Ema2 is required for DNA elimination, we analysed
exconjugants of an EMA2 somatic KO strain in which all MAC copies of
\emph{EMA2} were disrupted. DNA fluorescent in~situ hybridization (DNA-FISH)
with probes complementary to the transposable element Tlr1
\citep{cheng2010transposase} was performed at 36~hpm, a time by which DNA
elimination is normally completed \citep{chalker2011ciliate}. In wild-type
exconjugants, Tlr1 signal was observed only in the micronucleus (MIC),
consistent with complete elimination. In EMA2 KO exconjugants, Tlr1 signal
was detected in both the MIC and the new MAC, indicating retention of
IESs. Quantification of the IES retention index showed that the defect in
EMA2 KO cells was less severe than in TWI1 KO cells, where DNA elimination
is fully blocked \citep{mochizuki2002twi1}; the Welch two-sample $t$-test
gave a $p$-value of less than $0.001$ for the comparison with wild-type
exconjugants. Consistent with the coupling between DNA elimination and
progeny viability \citep{cheng2010transposase}, EMA2 KO cells failed to
produce viable sexual progeny in two independent crosses, whereas all
control wild-type crosses yielded viable progeny
(Figure~\ref{fig:phenotypes}). Ema2 is therefore required for completing
DNA elimination and for making viable sexual progeny.

\subsection{Ema2 is required for target-directed small RNA degradation}

We next asked whether Ema2 is involved in TDSD, the process by which
MAC-matching scnRNAs are selectively degraded in the parental MAC while
IES-matching scnRNAs are retained \citep{aronica2008ema1,schoeberl2012biased}.
Northern blot analysis using the 50-nucleotide Mi-9 probe, which recognizes
a repetitive sequence present in both IESs and MDSs, was performed on total
RNA collected at 3, 4.5 and 6~hpm from wild-type and EMA2 KO cells. In
wild-type cells, Mi-9-complementary scnRNAs were detected at 3~hpm, reduced
at 4.5~hpm, and became undetectable at 6~hpm, reflecting efficient TDSD
(Figure~\ref{fig:tdsd}). In EMA2 KO cells, however, Mi-9-complementary
scnRNAs remained detectable at 6~hpm, showing that Ema2 is required for
TDSD.

To extend this result genome-wide, we sequenced small RNAs at 3 and 8~hpm
from wild-type, EMA2 KO and EMA1 KO strains. Sequenced 26- to 32-nucleotide
reads, which correspond to scnRNAs \citep{schoeberl2012biased}, were mapped
to genomic tiles of MDSs and IESs. Statistical significance was assessed
using the Wilcoxon rank sum test, reported with three asterisks for
$p < 0.001$ and ``ns'' for $p > 0.05$. In wild-type cells, the
RPKM-normalized abundance of MDS-complementary scnRNAs was strongly reduced
from 3 to 8~hpm, as expected for TDSD, whereas in EMA2 KO cells the
reduction was only mild. At 3~hpm, MDS-complementary scnRNAs were
already more abundant in EMA2 KO than in wild-type cells, which is
consistent with scnRNA production and TDSD occurring concurrently. The
TDSD defect was milder in EMA2 KO cells than in EMA1 KO cells, in which
MDS-complementary scnRNAs remained constant between 3 and 8~hpm, indicating
that Ema1 and Ema2 act through different mechanisms. By contrast,
IES-complementary scnRNAs were comparable between 3 and 8~hpm across all
strains, consistent with the canonical view that TDSD targets only
MDS-complementary scnRNAs \citep{aronica2008ema1,schoeberl2012biased}.

\subsection{Ema2 is required for conjugation-stage accumulation of
SUMOylated proteins}

Inspection of the Ema2 primary sequence identified an SP-RING domain, a
signature of many SUMO E3 ligases \citep{hochstrasser2001spring}. The
Ema2 SP-RING domain is atypical: the first cysteine of the canonical
zinc-binding cluster is replaced by histidine, and several stabilizing
residues identified in Siz1 and Mms21 structures
\citep{yunus2009siz1,duan2009mms21} appear to be missing. Because SUMO
E3 ligases must engage the SUMO E2 conjugating enzyme Ubc9
\citep{sampson2001sumo}, we tested whether Ema2 binds Ubc9 directly.
We expressed \emph{E.~coli} codon-optimized GST and GST-Ema2, together
with His-Ubc9 (\emph{EMA2-Ec} and \emph{UBC9-Ec}), in \emph{E.~coli} BL21
and performed GST pull-downs. In contrast to GST alone, GST-Ema2 retained
His-Ubc9 on the glutathione beads, demonstrating a direct interaction
consistent with a bona fide SUMO E3 ligase activity.

To test whether Ema2 is responsible for protein SUMOylation during
conjugation, we expressed HA-tagged Smt3 (HA-Smt3) in an EMA2 KO strain.
Because SUMO is encoded by the single gene \emph{SMT3} in Tetrahymena
\citep{nasir2015sumo}, and because HA-Smt3 rescued the essential function
of Smt3 in SMT3 KO strains in a genetic rescue experiment, the HA-tagged
construct reports on bulk SUMOylation. The HA-Smt3-expressing EMA2 KO
strain was crossed with either a wild-type strain (``WT cross'') or
another EMA2 KO strain (``EMA2 KO cross''). Total proteins were harvested
at 4.5 and 6~hpm and analysed by anti-HA western blotting. At both time
points, high molecular weight HA-Smt3-conjugated proteins mainly greater
than 200~kDa were clearly detected in the WT cross, whereas their signal
was reduced to approximately 50\% in the EMA2 KO cross
(Figure~\ref{fig:sumo}). Normalizing the WT cross to one, three
independent experiments gave a WT-cross average of 1.0 and a KO-cross
average of approximately 0.5, with significance estimated by the Welch
two-sample $t$-test. These data indicate that Ema2 is the major SUMO E3
ligase during the mid-conjugation stage, while residual SUMOylation
presumably reflects other SP-RING-containing proteins
(TTHERM\_00227730, TTHERM\_00442270, TTHERM\_00348490) and/or
E3-independent SUMOylation \citep{sampson2001sumo}. A role for the SUMO
pathway in ciliate genome remodelling is consistent with earlier RNAi
experiments in \emph{Paramecium tetraurelia}, where knockdown of
\emph{UBA2} or \emph{SUMO} blocked IES excision \citep{matsuda2006sumo},
and with the essential function of SUMOylation in Tetrahymena cell
pairing \citep{nasir2015sumo}.

\subsection{Spt6 is the major Ema2-dependent SUMOylation target}

To identify Ema2-dependent substrates, we introduced a construct expressing
His-tagged Smt3 (His-Smt3) into wild-type and EMA2 KO backgrounds and
performed WT and EMA2 KO crosses. Total proteins were harvested at 6~hpm,
His-Smt3-conjugated proteins were enriched on Ni-NTA beads and identified
by mass spectrometry. To control for non-specific binders, we additionally
profiled wild-type cells without His-Smt3 expression and excluded any
protein whose $\log_2$ label-free quantification (LFQ) score exceeded 25
or that contained more than six consecutive histidine residues. While the
bulk of the identified proteins were similarly detected between the WT and
EMA2 KO crosses, Spt6 stood out as both the most abundantly recovered
protein in the WT cross and dramatically reduced in the EMA2 KO cross,
nominating it as the major Ema2 SUMOylation target.

We confirmed this assignment by expressing HA-tagged Spt6 (Spt6-HA) from
the endogenous \emph{SPT6} locus in an EMA2 KO background and crossing the
strain with either a wild-type partner or another EMA2 KO strain. Total
proteins harvested at 4.5 and 6~hpm were analysed by anti-HA western
blotting. In the WT cross, a slower-migrating Spt6-HA population appeared
in addition to the unmodified band at both time points, whereas in the
EMA2 KO cross this slower-migrating species was barely detectable.
A parallel analysis in which Spt6-HA was expressed in a wild-type
background confirmed that the slower-migrating forms are absent in
vegetatively growing cells and in starved cells at 0~hpm, that they
accumulate from the mid-conjugation stage, and that they are also present
at 8~hpm when the new MAC has formed, suggesting possible Spt6
SUMOylation in the new MAC as well. Immunoprecipitation of Spt6-HA with
an anti-HA antibody followed by anti-Smt3 blotting confirmed that the
slower-migrating species carry SUMO and are strongly reduced in the
EMA2 KO cross. Ema2 is therefore required for the majority of Spt6
SUMOylation during the mid-conjugation stage.

\subsection{Ema2 is required for pMAC-lncRNA accumulation and H3K27me3
deposition}

Because Spt6 is a conserved regulator of transcription elongation and TDSD
was proposed to be triggered by base pairing between scnRNAs and nascent
pMAC-lncRNAs \citep{aronica2008ema1,schoeberl2012biased}, we hypothesized
that Ema2 promotes pMAC-lncRNA transcription. We tested this by detecting
pMAC-lncRNAs spanning the M, L8 and R2 IES-MDS boundaries in conjugating
wild-type and EMA2 KO cells at 6~hpm using nested RT-PCR with the primer
sets listed in Supplementary Information. At all three loci, pMAC-lncRNAs
were detected in wild-type cells but not in EMA2 KO cells, while the
constitutively expressed RPL21 mRNA was detected in both conditions. As a
further control, genomic PCR with the same primer sets detected the
corresponding MAC loci in EMA2 KO cells, ruling out loss of primer binding
sites due to alternative DNA elimination in prior sexual generations. Ema2
is therefore required for the accumulation of pMAC-lncRNAs at least at
these three tested loci.

To probe the effect at a cytological level, we used the long double-stranded
RNA-specific J2 antibody to visualize dsRNAs formed by bidirectional
lncRNA transcription. In wild-type cells, long dsRNAs were first detected
in the MIC at 3~hpm, then in the parental MAC at 6~hpm, and finally in the
new MAC at 9~hpm. In EMA2 KO cells, long dsRNAs were detected in the MIC
at 3~hpm and in the new MAC at 9~hpm but were undetectable in the parental
MAC at 6~hpm, demonstrating that Ema2 is specifically required for
pMAC-lncRNA accumulation. In EMA1 KO cells, by contrast, parental-MAC
dsRNAs were still detected, consistent with the prior view that Ema1
promotes scnRNA--lncRNA interaction rather than lncRNA transcription
\citep{aronica2008ema1}.

We also examined H3K27me3, a scnRNA-dependent modification deposited in the
parental MAC by the histone methyltransferase Ezl1 within
PRC2 \citep{liu2007rnai,frapporti2019ezl1}. In conjugating cells at 6~hpm,
H3K27me3 was undetectable in the parental MAC in EMA2 KO cells and, as
expected from prior work, was also absent in EMA1, TWI1 (complete
somatic+germline) and EZL1 KO cells \citep{liu2007rnai}. Thus, Ema2 is
required for scnRNA-directed H3K27me3 deposition, likely as a consequence
of losing the chromatin recruitment of PRC2 by the scnRNA--lncRNA
interaction.

\subsection{Ema2 promotes chromatin association of Spt6 and RNA
polymerase~II}

We next asked how Ema2 influences Spt6 function. Immunofluorescence
staining of conjugating cells at 4.5~hpm showed that both Spt6-HA and
Rpb3, the third-largest subunit of RNA polymerase~II
\citep{mochizuki2004rpb3}, remained localized to the MAC in EMA2 KO
cross as well as in the WT cross, and also to the new MAC at 9~hpm. At
the cytological level, therefore, Ema2 is dispensable for the
macronuclear localization of Spt6 and RNAPII.

To probe subnuclear partitioning, we performed cell fractionation at
4.5~hpm. Conjugating cells were incubated with a lysis buffer that
releases cytoplasmic and nucleoplasmic proteins into the soluble fraction
(S1). The insoluble pellet was resuspended in fresh lysis buffer and
sonicated to release chromatin-bound proteins (S2). In the WT cross,
Spt6-HA and Rpb3 were detected in both the S1 and S2 fractions, and the
slower-migrating SUMOylated Spt6-HA species was restricted to the S2
(chromatin-bound) fraction. In the EMA2 KO cross, the amounts of
Spt6-HA and Rpb3 in the S2 fraction were greatly reduced. These findings
were reproduced in two additional biological replicates.

Because Twi1 binds chromatin via nascent pMAC-lncRNAs
\citep{aronica2008ema1}, we tested whether the chromatin association of
Spt6 and RNAPII depends on RNA. We treated the insoluble pellet with
RNase~A at 20~$\mu$g/mL before obtaining solubilized (S2) and insoluble
(I) fractions. Spt6-HA and Rpb3 remained in the insoluble fraction after
RNase~A treatment, whereas Twi1 was mostly released into the soluble
fraction, indicating that the chromatin association of Spt6 and RNAPII is
not RNA-dependent. We therefore conclude that Ema2 promotes the
association of Spt6 and RNAPII with parental-MAC chromatin at the
mid-conjugation stage.

\subsection{SUMOylation of Spt6 itself is dispensable for pMAC-lncRNA
transcription}

To directly test whether Spt6 SUMOylation is the functional consequence of
Ema2 activity, we generated a SUMOylation-defective Spt6 mutant.
Tetrahymena Spt6 consists of an N-terminal domain of low complexity (DOL)
followed by conserved TEX-like and SH2 domains. We produced HA-tagged Spt6
constructs in which lysine (K) residues were substituted by arginine (R)
either across all DOL lysines (HA-SPT6-DOL-KR) or across surface lysines
predicted by an AlphaFold model of Tetrahymena Spt6 in the N, M or C
non-DOL regions (HA-SPT6-N-KR, HA-SPT6-M-KR and HA-SPT6-C-KR,
respectively). Lysines predicted to be SUMOylation targets by the JASSA
algorithm in all three regions were included in the corresponding mutants.

All four mutants as well as the HA-SPT6-WT control rescued the vegetative
lethality of SPT6 KO cells, obtained through genetic rescue of SPT6
homozygous heterokaryons. However, cells rescued by HA-SPT6-N-KR and
HA-SPT6-M-KR showed severe defects in meiotic progression and mating
initiation, respectively, which precluded assessment of their SUMOylation
status during conjugation. Cells rescued by HA-SPT6-DOL-KR and
HA-SPT6-C-KR progressed through conjugation, with HA-SPT6-C-KR showing
a lower mating efficiency that we compensated for by loading three times
more total protein in western blot experiments.

Anti-HA western blotting showed that HA-Spt6-DOL-KR was SUMOylated like
wild-type HA-Spt6, producing both unmodified (UM) and modified (M) bands,
whereas HA-Spt6-C-KR yielded no detectable slower-migrating SUMOylated
species even when three times more material was loaded. Spt6-C-KR
therefore represents a SUMOylation-defective Spt6 mutant, with at least a
further reduction of SUMOylation compared with the level observed for
Spt6 in the absence of Ema2.

Despite losing detectable SUMOylation, HA-SPT6-C-KR cells retained Spt6 and
RNAPII in the chromatin-bound S2 fraction across three independent cell
fractionation experiments, accumulated long dsRNAs in the parental MAC as
detected by J2 immunostaining at 4.5~hpm, and also showed long dsRNAs in
the new MAC. Nevertheless, exconjugants from HA-SPT6-C-KR cells analysed
at 36~hpm by Tlr1 DNA-FISH showed a mild DNA elimination defect. Spt6
SUMOylation per se is therefore dispensable for pMAC-lncRNA transcription,
but still contributes to efficient DNA elimination, implying that Ema2
acts through additional SUMOylated substrates during DNA elimination
(Figure~\ref{fig:spt6mutants}).

\section{Discussion}

Our experiments identify Ema2 as the first factor required for genome-wide
pMAC-lncRNA transcription in \emph{Tetrahymena thermophila}. In the
absence of Ema2, pMAC-lncRNAs and parental-MAC dsRNAs fail to accumulate
at 6~hpm, Mi-9-complementary scnRNAs remain undegraded at 6~hpm and are
strongly enriched in MDS tiles at 8~hpm, H3K27me3 is lost from the parental
MAC at 6~hpm, DNA elimination is partially blocked, and sexual progeny are
inviable. All of these defects can be rationalized by a single upstream
event: the failure to transcribe pMAC-lncRNAs, which normally provide the
scaffold for scnRNA base pairing and downstream PRC2 recruitment. In this
model, Ema2-directed SUMOylation acts as a licence that makes parental-MAC
chromatin transcriptionally competent for lncRNA production.

A striking observation is that Ema2 enhances the chromatin interaction of
Spt6 and RNA polymerase~II specifically during conjugation, whereas Spt6
and RNAPII are essential for vegetative cell viability
\citep{mochizuki2004rpb3}. Because EMA2 KO cells proceed through most of
conjugation without obvious arrest, essential mRNA transcription must
still occur in the parental MAC in their absence, even while the bulk of
Spt6 and RNAPII is lost from chromatin. The most economical interpretation
is that the mid-conjugation loss of chromatin Spt6/RNAPII is transient and
specifically affects pMAC-lncRNA transcription: a radical change in the
parental-MAC chromatin environment could destabilize the transcription
machinery, and Ema2-dependent SUMOylation might maintain Spt6 and RNAPII
on chromatin within that window. Alternatively, because many transcriptional
regulators are strongly upregulated during conjugation
\citep{miao2009microarray}, pMAC-lncRNA transcription may require a
quantitative excess of Spt6 and RNAPII beyond the housekeeping pool, and
Ema2 could be needed to commit this excess to chromatin.

Unexpectedly, the SUMOylation-defective Spt6-C-KR mutant preserves
chromatin-bound Spt6 and RNAPII and supports pMAC-lncRNA accumulation,
even though it is refractory to SUMOylation at the resolution of our
western blots. This result argues that SUMOylation of Spt6 itself is not
the relevant Ema2 output for pMAC-lncRNA transcription. Several non-mutually
exclusive explanations remain open. Other Ema2-dependent substrates may
have escaped detection in our mass spectrometry screen, for example because
Ema2-dependent SUMOylation can compete with, or drive the turnover of, the
same proteins via SUMO-targeted ubiquitination
\citep{praefcke2012sumo}. Alternatively, SUMOylation on Spt6 may compete
with other modifications on the same lysine residues, and the Spt6-C-KR
mutant may inadvertently mimic a constitutively SUMOylated state by
blocking those competing modifications. Identifying individual
Ema2-dependent SUMOylated lysines in vivo, ideally under proteasome
inhibition, and profiling other post-translational modifications on Spt6
in the presence and absence of Ema2, will be needed to distinguish between
these possibilities. These analyses are also the most direct route to
understanding the mild DNA elimination defect of the Spt6-C-KR mutant
itself.

SUMOylation has been repeatedly linked to small RNA-directed silencing.
In fission yeast, SUMOylation of Swi6, Chp2 and Clr4 supports heterochromatin
maintenance \citep{shin2005sumo}. In \emph{C.~elegans}, SUMOylation of
HDAC1 promotes Argonaute-directed transcriptional silencing
\citep{kim2021hdac1}. In Drosophila, the SUMO E3 ligase Su(var)2-10
couples piRNA target recognition to SetDB1-driven H3K9me3 deposition, and
an Su(var)2-10-independent SUMOylation of Panoramix engages the general
heterochromatin effector Sov \citep{ninova2019suvar}. In each of these
cases SUMOylation modifies factors that act on pre-existing lncRNAs. Our
work establishes that in Tetrahymena the Ema2-directed SUMO pathway acts
\emph{upstream} of lncRNA transcription itself. Furthermore, loss of the
Piwi protein Twi1 is known to block H3K27me3 in the parental MAC without
affecting lncRNA accumulation \citep{liu2007rnai,mochizuki2002twi1},
showing that heterochromatin is not a prerequisite for pMAC-lncRNA
transcription in Tetrahymena, in contrast to heterochromatin-dependent
lncRNA transcription in Drosophila piRNA clusters
\citep{andersen2017heterochromatin}. Together with the essential role of
the SUMO pathway for IES excision in \emph{Paramecium}
\citep{matsuda2006sumo}, these observations place ciliate SUMO E3 ligases
at a mechanistically distinctive node: directly enabling lncRNA
transcription that in turn feeds small RNA-directed genome surveillance
(Figure~\ref{fig:pathway}).

We summarize the categorical phenotypes observed across the strains in
this study in Figure~\ref{fig:phenotypes} and
Tables~\ref{tab:timeline}--\ref{tab:spt6mutants} below, and we anticipate
that further dissection of Ema2-dependent SUMO targets will reveal
additional regulatory layers that link post-translational modification to
transcriptional licensing during programmed DNA elimination.

\begin{table}[htbp]
\centering
\caption{Ema2 expression and localization timeline during conjugation.}
\label{tab:timeline}
\begin{tabular}{llll}
\toprule
Stage & Time (hpm) & EMA2 mRNA & Ema2-HA localization \\
\midrule
Vegetative      & --    & not expressed           & not detectable              \\
Early conj.     & 3     & expressed (conjugation) & parental MAC                \\
Mid conj.       & 6     & expressed (conjugation) & parental MAC                \\
New MAC onset   & 8     & expressed (conjugation) & new MAC                     \\
Late conj.      & 12    & expressed (conjugation) & fading (new MAC)            \\
Late conj.      & 14    & expressed (conjugation) & fading (new MAC)            \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\caption{Phenotypes of EMA2 somatic KO compared with related knockouts.
``+'' denotes the feature is present/complete in wild type, ``--'' denotes
absent/defective, ``partial'' denotes an intermediate phenotype, and
``n.d.'' denotes features not explicitly measured for that strain in the
experimental log.}
\label{tab:phenotypes}
\begin{tabular}{llllll}
\toprule
Strain    & DNA elim.     & MDS scnRNA  & pMAC-lncRNA / & H3K27me3  & Viable           \\
          & (Tlr1 FISH)   & TDSD        & dsRNA         & in pMAC   & progeny          \\
\midrule
WT        & +             & +           & +             & +         & +                \\
EMA2 KO   & partial       & --          & --            & --        & --               \\
EMA1 KO   & --            & -- (severe) & +             & --        & n.d.             \\
TWI1 KO   & --            & n.d.        & n.d.          & --        & n.d.             \\
EZL1 KO   & n.d.          & n.d.        & n.d.          & --        & n.d.             \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\caption{Summary of the Spt6 K-to-R mutant series. Entries marked ``n.d.''
were not examined in the experimental log for that mutant.}
\label{tab:spt6mutants}
\begin{tabular}{llllll}
\toprule
Mutant      & Substitutions        & Veg. rescue & Conj. phenotype       & SUMOylation & DNA elim.   \\
\midrule
HA-SPT6-WT     & none                 & yes         & normal                & detected    & +           \\
HA-SPT6-DOL-KR & all DOL lysines      & yes         & normal                & detected    & n.d.        \\
HA-SPT6-N-KR   & N-region surface K   & yes         & meiotic block         & n.d.        & n.d.        \\
HA-SPT6-M-KR   & M-region surface K   & yes         & mating block          & n.d.        & n.d.        \\
HA-SPT6-C-KR   & C-region surface K   & yes         & lower mating, progresses & not detected & mild defect \\
\bottomrule
\end{tabular}
\end{table}

\input{figures/figures.tex}

\section{Materials and Methods}

\paragraph{Strains and cell culture.}
Wild-type \emph{Tetrahymena thermophila} strains B2086, CU427 and CU428,
and the MIC-defective ``star'' strains B*VI and B*VII, were obtained from
the Tetrahymena stock center. EMA2 somatic KO strains were described
previously (see Supplementary Information for strain details). Cells were
grown in SPP medium containing 2\% proteose peptone at $30\,^{\circ}$C.
For conjugation, growing cells (${\sim}5$--$7 \times 10^{5}$\,cells/mL)
of two different mating types were washed, pre-starved for
${\sim}12$--$24$~hours, and mixed in 10~mM Tris pH 7.5 at $30\,^{\circ}$C.

\paragraph{Antibodies.}
The following antibodies were obtained from commercial sources:
anti-H3K27me3 (Millipore, 07-449); anti-HA HA11 (Covance, clone 6B12);
anti-H3K9me3 (Active Motif, 39162); anti-GST (BD Biosciences, 554805);
anti-His (Proteintech, 66005-1-Ig); anti-long dsRNA J2 (Jena Bioscience,
RNT-SCI-10010200); anti-$\alpha$-tubulin 12G10 (Developmental Studies
Hybridoma Bank). The anti-Rpb3 antibody was as described previously
\citep{mochizuki2004rpb3}, and the anti-Smt3 antibody was a gift from
Dr.~James Forney \citep{nasir2015sumo}. All primary and secondary
antibodies were diluted 1:1{,}000 for immunofluorescence and 1:10{,}000
for western blotting, with HRP-coupled secondaries used for western
blotting.

\paragraph{DNA elimination assay by FISH.}
DNA-FISH was performed using Cy5-labelled probes generated by nick
translation from the plasmids pMBR~4C1, pMBR~2 and Tlr~IntB, which cover
different parts of the Tlr1 element \citep{cheng2010transposase}. Images
of stained cells were analysed with Fiji, and regions corresponding to
the new MAC (A), MIC (I) and a cytoplasmic area (C) were manually
outlined. The IES retention index was computed as
\[
\text{IES retention} \;=\;
0.35 \times \frac{(\bar{\text{Cy3}}_A - \bar{\text{Cy3}}_C) /
(\bar{\text{Cy3}}_I - \bar{\text{Cy3}}_C)}
{(\bar{\text{DAPI}}_A - \bar{\text{DAPI}}_C) /
(\bar{\text{DAPI}}_I - \bar{\text{DAPI}}_C)} ,
\]
where the correction factor of 0.35 was determined empirically so that the
average IES retention index of TWI1 KO cells equals 1, under the
assumption that DNA elimination is completely blocked in TWI1 KO cells.
Three asterisks indicate $p < 0.001$ in a Welch two-sample $t$-test.

\paragraph{Viability test for sexual progeny.}
Mating pairs were isolated into drops of $1\times$SPP at 6~hpm and
incubated for ${\sim}3$~days. Progeny viability was assessed by 6-methylpurine
(6-mp) resistance and paromomycin sensitivity, following established
procedures for Tetrahymena conjugation.

\paragraph{Small RNA analysis.}
For northern blots, 10~$\mu$g of total RNA isolated from conjugating
wild-type and EMA2 KO cells (with TRIzol) was separated on a 15\%
acrylamide-urea gel and hybridized with the radiolabelled 50-nt Mi-9
probe, which is complementary to a repetitive MDS sequence
\citep{aronica2008ema1}. For small RNA sequencing, 26- to 32-nucleotide
reads corresponding to scnRNAs \citep{schoeberl2012biased} that uniquely
matched MDS or IES genomic tiles were counted and normalized as
reads per kilobase of unique sequence per million (RPKM). Significance was
assessed using the Wilcoxon rank sum test.

\paragraph{GST pull-down assay.}
\emph{E.~coli} codon-optimized \emph{EMA2} (EMA2-Ec) and \emph{UBC9}
(UBC9-Ec) synthetic genes were cloned into the BamHI/XhoI sites of
pGEX-4T-1 and the NdeI/BamHI sites of pET28a(+) to generate
pGEX-4T-1-EMA2-Ec and pET28-UBC9-Ec, respectively. His-Ubc9 was expressed
in BL21 \emph{E.~coli} by induction with 0.05~mM
isopropylthio-$\beta$-galactoside (IPTG) for ${\sim}16$~hours at
$25\,^{\circ}$C. Cells were lysed in HEPES buffer (50~mM HEPES pH 7.5,
800~mM KCl, 10\% glycerol, 0.2~mM PMSF, 1$\times$ cOmplete EDTA-free
protease inhibitors) containing 10~mM imidazole, and the lysate was
incubated with Ni-NTA beads for 2~hours at $4\,^{\circ}$C. After four
washes in HEPES buffer with 20~mM imidazole, His-Ubc9 was eluted with
HEPES buffer containing 250~mM imidazole, dialysed against 50~mM Tris
pH 7.5, 150~mM NaCl, 2~mM MgCl$_2$, 1~mM DTT, 50\% glycerol, and passed
through a 10~kDa Microcon filter. For the pull-down, BL21 cells expressing
GST or GST-Ema2 from pGEX-4T-1 or pGEX-4T-1-EMA2-Ec were lysed in 50~mM
Tris-HCl pH 7.5, 100~mM NaCl, 1~mM DTT, 0.2~mM PMSF, 1$\times$ cOmplete
EDTA-free, and incubated with Glutathione Sepharose 4B for 90~minutes at
$4\,^{\circ}$C. 10~$\mu$g of His-Ubc9 was then added in GST pull-down
buffer (20~mM Tris-HCl pH 7.5, 100~mM NaCl, 1\% NP-40, 5~mM MgCl$_2$,
1~mM DTT, 0.2~mM PMSF, 1$\times$ cOmplete EDTA-free) and incubated for
90~minutes at $4\,^{\circ}$C. Beads were washed four times in GST
pull-down buffer and eluted with 2$\times$ SDS sample buffer.

\paragraph{Purification of His-Smt3-conjugated proteins.}
Cells were lysed in guanidine lysis buffer (6~M guanidine-HCl, 93.2~mM
Na$_2$HPO$_4$, 6.8~mM NaH$_2$PO$_4$, 10~mM Tris-HCl pH 8.0) by
sonication (2$\times$25 pulses at 20\% power/60\% pulse using an
Omni-Ruptor~250 homogenizer). The cleared lysate was supplemented with
50~mM imidazole and 5~mM $\beta$-mercaptoethanol and incubated with
Ni-NTA agarose beads. After washing, bound proteins were eluted in
elution buffer (6~M urea, 58~mM Na$_2$HPO$_4$, 42~mM NaH$_2$PO$_4$,
10~mM Tris-HCl pH 8.0, 500~mM imidazole), precipitated in 10\%
trichloroacetic acid (TCA), resuspended in 1$\times$ SDS buffer, and
analysed by western blotting and mass spectrometry.

\paragraph{Purification of HA-Smt3-conjugated proteins.}
Total proteins were precipitated by incubating cells in 6\% TCA on ice
for 10~minutes, centrifuged at 13{,}000~rpm for 5~minutes at
$4\,^{\circ}$C, dissolved in 1$\times$ SDS sample buffer, neutralized
with 2~M Tris, and heated at $95\,^{\circ}$C for 5~minutes. 600~$\mu$L
of the dissolved proteins was then diluted in 7.4~mL of 20~mM Tris
pH 7.5, 100~mM NaCl, 2~mM MgCl$_2$, 2~mM CaCl$_2$ and incubated overnight
at $4\,^{\circ}$C with EZview~Red anti-HA agarose beads (Sigma). Beads
were washed five times in 20~mM Tris pH 7.5, 100~mM NaCl, 2~mM MgCl$_2$,
2~mM CaCl$_2$, 0.1\% Tween-20, and the bound proteins were eluted in
1$\times$ SDS buffer at $95\,^{\circ}$C for 5~minutes.

\paragraph{Cell fractionation.}
Cell fractionation was performed as previously described. Conjugating
cells ($1.4\times10^{6}$) were incubated for 30~minutes on ice with
1~mL of lysis buffer (50~mM Tris pH 8.0, 100~mM NaCl, 5~mM MgCl$_2$,
1~mM EDTA, 0.5\% Triton~X-100, 1$\times$ cOmplete protease inhibitor)
and centrifuged at 15{,}000$g$ for 10~minutes at $4\,^{\circ}$C to
separate soluble (S1) and insoluble fractions. Proteins in 500~$\mu$L
of supernatant were precipitated with 10\% TCA, washed three times with
ice-cold acetone, and dissolved in 70~$\mu$L of 1$\times$SDS sample
buffer. The insoluble fraction was resuspended in 2~mL of fresh lysis
buffer, sonicated (5 bursts of 20\% power, 50\% pulse, 5 pulses each
with $>20$~seconds interval) and centrifuged at 20{,}000$g$ for
10~minutes at $4\,^{\circ}$C; the resulting supernatant (solubilized
chromatin, S2) was TCA-precipitated and dissolved as above. For RNase~A
treatment, the insoluble fraction was first incubated with 20~$\mu$g/mL
RNase~A in fresh lysis buffer before the S2 and insoluble (I) fractions
were obtained.

\paragraph{RT-PCR analyses of lncRNA transcripts.}
Total RNA from mating wild-type and EMA2 KO cells at 6~hpm was isolated
with TRIzol. lncRNA transcripts spanning the M, L8 and R2 IES boundaries
were amplified by nested PCR ($95\,^{\circ}$C for 20~s / $50\,^{\circ}$C
for 30~s / $68\,^{\circ}$C for 1~min, 30 cycles each) with the primer sets
M5$'$-3 + M3$'$-3 (M-1st), M5$'$-4 + M5$'$-4 (M-2nd), L8\_5$'$-1 + L8\_3$'$-1
(L8-1st), L8\_5$'$-2 + L8\_3$'$-2 (L8-2nd), R2\_5$'$-1 + R2\_3$'$-1 (R2-1st)
and R2\_5$'$-2 + R2\_3$'$-2 (R2-2nd), and analysed on 1\% agarose gels.
RPL21 mRNA was amplified with RPL21-FW and RPL21-RV as a positive control.

\paragraph{SPT6 germline knockout strains.}
To construct the knockout, the 5$'$ and 3$'$ flanking regions of the
\emph{SPT6} gene were amplified by PCR with SPT6-KO-5F-FW/5F-RV and
SPT6-KO-3F-FW/3F-RV primer pairs and concatenated by overlap PCR using
SPT6-KO-5F-FW and SPT6-KO-3F-RV. Transformed cells were selected in
300~$\mu$g/mL puromycin and 1~$\mu$g/mL CdCl$_2$ and phenotypically
assorted until cells grew in 1.2~mg/mL of puromycin.

\paragraph{Spt6 lysine-to-arginine mutants.}
Surface lysines in the N, M and C non-DOL regions of an AlphaFold-predicted
Tetrahymena Spt6 structure were substituted by arginine. Residues also
predicted to be SUMOylation targets by the JASSA algorithm in all three
non-DOL regions were additionally included in HA-SPT6-N-KR, HA-SPT6-M-KR
and HA-SPT6-C-KR constructs. HA-SPT6-DOL-KR carried K-to-R substitutions
for all lysines within the N-terminal domain of low complexity. Constructs
were introduced into SPT6 KO cells and assayed for their ability to rescue
vegetative lethality.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
