\section{Introduction}

A phylogenetic tree depicts the evolutionary history of a given set of species
and efficient, accurate estimation of phylogenies from genome data remains a
central problem in bioinformatics and systematics. Traditional tree-inference
approaches fall into two broad groups. Distance-based methods, such as UPGMA
\citep{sokal1958upgma} and neighbor-joining \citep{saitou1987nj}, first
summarize every pair of input sequences as a scalar distance and then apply an
agglomerative algorithm to the resulting matrix. Character-based methods, such
as maximum parsimony \citep{fitch1971parsimony} and maximum likelihood
\citep{felsenstein1981ml, posada2021felsenstein}, retain the per-site character
information throughout inference and score candidate trees under an explicit
model of substitution. Within the alignment-based paradigm, maximum likelihood
has repeatedly been shown to be more accurate and more robust than
distance-based methods, particularly when the assumptions of the distance
model are violated \citep{felsenstein1981ml, posada2021felsenstein}.

Both character- and distance-based approaches in the alignment-based paradigm
rely on a multiple sequence alignment (MSA) of the input sequences. The
quality of that alignment directly governs the quality of the inferred tree,
but constructing an optimal MSA is computationally intractable in the general
case \citep{elias2006msa}, and even heuristic multi-genome alignment is
memory- and time-intensive for genome-scale inputs. A further, biological
limitation is that alignment assumes a preserved linear order of homology, so
that rearrangement events such as translocations and inversions inside
whole-genome sequences complicate alignment and can degrade downstream
phylogenetic inference.

To circumvent these limitations, a growing body of work has turned to
\emph{alignment-free} methods, which replace MSA with sequence summaries such
as $k$-mer frequency vectors or MinHash sketches and derive a distance
directly from these summaries \citep{vinga2003review, zielezinski2017review}.
Alignment-free methods scale to whole genomes and, unlike assembly-based
alignment pipelines, apply to genome skims in which the nuclear genome is not
assembled. The community benchmarking effort of the AFproject
\citep{zielezinski2019afproject} has compared dozens of such methods and
documented a persistent accuracy gap between alignment-free and alignment-based
approaches, with no single alignment-free method dominating across the
benchmark's heterogeneous datasets.

Despite the clear methodological parallel between alignment-based inference
(where maximum likelihood outperforms distance-based inference) and
alignment-free inference (which remains almost entirely distance-based), no
widely used maximum-likelihood alignment-free phylogeny tool exists. The
pattern-based Bayesian approach of \citet{hohl2007alignmentfree} is the main
prior attempt to bring likelihood machinery to alignment-free inference, but
it was shown not to improve on classical distance-based alignment-free
methods in their own benchmarks, and no established software pipeline has
emerged. More recent spaced-word-based work \citep{dencker2020multispam}
incorporates quartet-level maximum likelihood on gap-free micro-alignments,
which is an important step but still differs from a full binary-trait
likelihood model over all shared $k$-mers.

In this paper, we present Peafowl (Phylogeny Estimation through Alignment
Free Optimization With Likelihood), the first alignment-free tool that
estimates a phylogenetic tree by running a standard nucleotide-scale
maximum-likelihood engine directly on a binary, $k$-mer presence/absence
matrix of the input species. The contributions of this work are:
\begin{itemize}
\item a likelihood-based alignment-free framework that encodes each species
as a binary vector of $k$-mer presence/absence and estimates tree topology
under the two-state BINGAMMA substitution model of RAxML
\citep{stamatakis2014raxml};
\item a data-driven, entropy-based procedure for selecting the $k$-mer
length by maximizing cumulative Shannon entropy \citep{shannon1948mathematical}
over a fixed number of sampled matrix rows across odd $k$ from 9 to 31;
\item an empirical evaluation on seven real phylogenomic datasets, including
five AFproject benchmarks, comparing Peafowl against nine state-of-the-art
alignment-free tools;
\item the observation that non-canonical $k$-mer counting recovers the
reference tree on an inversion-dominated Yersinia benchmark on which most
tools fail, demonstrating a principled way to expose rearrangement signal.
\end{itemize}

\section{Related Work}

\paragraph{$k$-mer based distance methods.}
The earliest alignment-free methods compared sequences by summarizing them
with $k$-mer frequency vectors and computing vector distances, an idea that
underlies feature frequency profiles (FFP) \citep{sims2009ffp}, composition
vectors (CVTree / CAFE-cvtree) \citep{gao2007cvtree, lu2017cafe}, and
general-purpose $k$-mer counting infrastructure such as Jellyfish
\citep{marcais2011jellyfish}. MinHash-based sketches later reduced the
storage and runtime cost of such comparisons: mash \citep{ondov2016mash}
uses a small MinHash sketch to estimate a mutation distance, and phylonium
\citep{klotzl2020phylonium} extends this idea by indexing a single
representative genome to produce accurate distances on very large samples of
closely related genomes.

\paragraph{Match-anchor and spaced-word methods.}
A second family of methods identifies conserved $k$-mer-like anchors and
derives evolutionary distances from them. The tool andi
\citep{haubold2015andi} searches for pairs of maximal unique matches of a
minimum length and uses the short ungapped regions between them as
micro-alignments. Co-phylog \citep{yi2013cophylog} uses fixed-length
``context'' micro-alignments with a single middle-mismatch position to
recover phylogenies from assembled or unassembled sequences. Filtered
Spaced-Word Matches (FSWM) \citep{morgenstern2015spacedwords}, its
low-coverage variant Read-SpaM \citep{lau2019readspam}, Slope-SpaM
\citep{rohling2020slopespam}, and Prot-SpaM \citep{leimeister2019protspam}
use carefully chosen patterns of match and don't-care positions to improve
the statistical reliability of inferred distances over contiguous $k$-mers.

\paragraph{Likelihood in alignment-free inference.}
Attempts to use likelihood ideas without a full MSA are comparatively rare.
\citet{hohl2007alignmentfree} introduced a Bayesian formulation over $k$-mer
presence on a small benchmark but did not outperform classical distance-based
alignment-free methods. Multi-SpaM \citep{dencker2020multispam} computes
maximum-likelihood quartet trees on gap-free micro-alignments and combines
them via Quartet MaxCut, which introduces likelihood at the quartet level
but still relies on spaced-word micro-alignments rather than on a global
character-by-character substitution model. Peafowl occupies a
previously empty niche: it runs a standard ML engine
\citep{stamatakis2014raxml} directly on a binary $k$-mer presence/absence
matrix, obtaining tree topologies without any alignment step at all.

\paragraph{Genome skims and assembly-free settings.}
Skmer \citep{sarmashghi2019skmer} extends alignment-free phylogenetic
placement to unassembled, low-coverage genome skims and has been adopted in
several taxonomic settings. Read-SpaM \citep{lau2019readspam} and co-phylog
\citep{yi2013cophylog} similarly support unassembled reads. Peafowl in its
current form operates on assembled genomes and so addresses a complementary
but overlapping regime; extension to unassembled reads is a natural next
step.

\paragraph{Benchmarking.}
We use the AFproject benchmark \citep{zielezinski2019afproject}, which
exercises alignment-free methods on five real-data categories including
genome-based phylogeny and horizontal gene transfer. Previous analyses
using this benchmark have observed that no single alignment-free method
wins on all datasets, and that inversion- and rearrangement-heavy datasets
(e.g., Yersinia) are particularly challenging. Tree comparisons throughout
this paper use the normalized Robinson--Foulds distance
\citep{robinson1981rf}, computed with PHYLIP
\citep{felsenstein1989phylip}; distance-only competitors are post-processed
into trees with neighbor-joining \citep{saitou1987nj} and UPGMA
\citep{sokal1958upgma} in MEGA X \citep{kumar2018megax}.
